Type-safe client-server communication for functional web applications

### 2014-02-28 paper.tex
Finish explanation of App monad and its operations.

\title{Type-safe client-server communication for web applications}
\subtitle{Subtitle Text, if any}

@@

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the aptly
named \lstinline!Serializable! type class, and returns a typed ``handle'' of
sorts. In this example, the type of \lstinline!greetings! is
\lstinline!Export (String -> Server ())!, indicating that the handle identifies
an exported function with a single \lstinline!String! argument and no return
value; exported functions all live in the \lstinline!Server! monad.

@@

@@

\begin{code}
type Method = [JSON] -> IO JSON
type AppState = (CallID, [(CallID, Method)])
newtype App a = App (StateT AppState IO a)
\end{code}

As we can see, \lstinline!App! is a simple state monad, with underlying IO
capabilities to allow server-side computations to be forked from within it.
Aside from the \lstinline!runClient! function, which is only used to launch the
client program, This monad has two primary operations.

\begin{code}
liftServerIO :: IO a -> App (Useless a)
export :: Exportable a => a -> App (Export a)
\end{code}

Aside from the \lstinline!App! monad itself, these functions operate on two
types we have not seen before.

\begin{code}
data Useless a = Useful a | Useless
data Export a = Export CallID [JSON]
\end{code}

@@

@@

The core of this interface consists of the \lstinline!ffi! function, which
allows the programmer to create a Haskell function from arbitrary Javascript
code. This function exploits Javascript's ability to parse and execute
arbitrary strings at run-time using the \lstinline!eval! function, coupled with
the fact that functions in Haste and in Javascript share the same
representation, to dynamically create a function object at runtime.
The \lstinline!ffi! function is typed using a well-known type class trick for
creating statically typed variadic functions\ \cite{printf}, and works very
much like the \lstinline!printf! function of Haskell's standard library.
When applied to one or more arguments instantiating the \lstinline!Marshal!
type class, the \lstinline!pack! function is applied to each argument,
marshalling them into their respective Javascript representations, before they
are passed to the dynamically created function. When that function returns,
the inverse \lstinline!unpack! function is applied to its return value before
it is passed back into Haskell land.

>>>>

\title{Type-safe client-server communication for functional web applications}

@@

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the aptly
named \lstinline!Serialize! type class, and returns a typed ``handle'' of
sorts. In this example, the type of \lstinline!greetings! is
\lstinline!Export (String -> Server ())!, indicating that the handle identifies
an exported function with a single \lstinline!String! argument and no return
value; exported functions all live in the \lstinline!Server! monad.

@@

To pass data back and forth over the network, messages are serialized using
JSON, a fairly lightweight format used by many web applications, and sent using
the HTML5 WebSockets API. This choice is completely arbitrary, guided purely
by implementation convenience. It is certainly not the most performant choice,
but can be easily replaced with something more suitable as the need arises.

@@

\begin{code}
type CallID = Int
type Method = [JSON] -> IO JSON
type AppState = (CallID, [(CallID, Method)])
type App = StateT AppState IO
\end{code}

As we can see, \lstinline!App! is a simple state monad, with underlying IO
capabilities to allow server-side computations to be forked from within it.
Its \lstinline!CallID! state element contains the identifier to be given to the
next exported function, and its other state element contains a mapping from
identifiers to exported functions. Of course, in reality, \lstinline!App! is a
\lstinline!newtype!, to prevent arbitrary lifting of IO computations.

Aside from the \lstinline!runClient! function, which is only used to launch the
client program, This monad has two primary operations.

\begin{code}
data Useless a = Useful a | Useless
data Export a = Export CallID [JSON]

liftServerIO :: IO a -> App (Useless a)
export :: Exportable a => a -> App (Export a)
\end{code}

@@

\lstinline!export! is a more interesting function. Looking at its type, we see
that it accepts any \lstinline!Exportable! value for its argument.
\lstinline!Exportable! is defined as follows.

\begin{code}
class Exportable a where
  mkExported :: a -> ([JSON] -> IO JSON)

instance Serialize a =>
         Exportable (Server a)
instance (Serialize a, Exportable b) =>
         Exportable (a -> b)
\end{code}

In essence, any function, over any number of arguments, which returns a
serializable value in the \lstinline!Server! monad can be exported. The
\lstinline!mkExported! function makes use a well-known type class trick for
creating statically typed variadic functions, and works very much like the
\lstinline!printf! function of Haskell's standard library.\ \cite{printf}
From an \emph{n}-ary function, it creates a unary function which expects
a list of \emph{n} JSON-encoded elements as its argument. When applied, its
argument list is deserialized and passed to the exported function, and its
return value is serialized. This allows us to easily bundle up all of an
exported function's arguments and pass them across the network.

The client implementation of \lstinline!export! is straightforward.

\begin{code}
export _ = do
  (next_id, exports) <- get
  put (next_id+1, exports)
  return (Export next_id [])
\end{code}

As the exported function lives on the server, the client only needs an
identifier to be able to call on it. The exported function is thus ignored,
so that it can be optimized out of existence in the client version of the code.
The server version is only marginally more complex in that it updates the
identifier-function mapping table as well.

\begin{code}
export f = do
  (next_id, exports) <- get
  put (next_id+1, (next_id, mkExported f) : exports)
  return (Export next_id [])
\end{code}

The second argument of the \lstinline!Export! constructor is, as the observant
reader might have guessed, the list of arguments to be sent to the export.
Essentially, an explicit closure is built up using the \lstinline!<.>! operator
and sent off over the network for execution.

\begin{code}
(<.>) :: Serialize a
      => Export (a -> b) -> a -> Export b
(Export identifier args) <.> arg =
  Export identifier (serialize arg : args)
\end{code}

@@

The core of this interface consists of the \lstinline!ffi! function, which
allows the programmer to create a Haskell function from arbitrary Javascript
code. This function exploits Javascript's ability to parse and execute
arbitrary strings at run-time using the \lstinline!eval! function, coupled with
the fact that functions in Haste and in Javascript share the same
representation, to dynamically create a function object at runtime.
The \lstinline!ffi! function is typed using the same method as the
\lstinline!serializify! function described in section \ref{sec:impl}.
When applied to one or more arguments instantiating the \lstinline!Marshal!
type class, the \lstinline!pack! function is applied to each argument,
marshalling them into their respective Javascript representations, before they
are passed to the dynamically created function. When that function returns,
the inverse \lstinline!unpack! function is applied to its return value before
it is passed back into Haskell land.
